On multilevel constraint decomposition methods \\ for nonlinear variational inequalities

### 2021-07-13 paper/mcd2.tex
Edits section 1.

The goal of this paper is to generalize the constraint decomposition (CD) method of X.-C.~Tai \cite{Tai2003}.  We extend this solution method to nonlinear variational inequality (VI) problems for which it has not been previously considered, we address important finite element (FE) implementation details, and we demonstrate the resulting performance.

The convergence of the CD method for several types of decompositions was proven in \cite{Tai2003} in the case of coercive VIs which arise from minimization of a convex functional over a convex set.  In its multilevel form the CD method has been shown to have optimal complexity for elliptic, linear obstacle problems \cite[subsection 5.4]{Tai2003}; see also Theorem 4.6 and Algorithm 4.7 in \cite{GraeserKornhuber2009}.

We extend the original algorithms, theory, and examples presented in \cite{Tai2003} in four particular directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item In contrast to \cite{Tai2003}, we do not assume that the continuum VI problem arises from optimization of a scalar objective.  Nonetheless we can prove convergence at the same rate (Section \ref{sec:convergence}); % HOPE
examples are explored in Sections \ref{sec:vi} and \ref{sec:results}.
\item We observe that the iterates from the CD method are always admissible, and thus the residual functional need only be defined on the admissible subset; Section \ref{sec:results} includes an example.  Admissible-iterate methods allow direct solutions of certain VI problems, such as fluid-layer dynamics problems \cite{Bueler2021conservation,JouvetBueler2012} for which non-admissible methods, such as semi-smooth methods \cite{BensonMunson2006}, may require unnatural modifications of the residual form.
\item In the multilevel case we make the observation that, if implemented appropriately, V(0,$\nu$) ``up-slash'' cycles are intrinsically more efficient than the V($\nu$,0) ``down-slash'' cycles (Section \ref{sec:multilevel}).  This observation seems to be new for multilevel CD methods; compare the comments on V(1,0) and V(1,1) cycles in \cite{GraeserKornhuber2009,Tai2003}.  Strong preference for up-smoothing is, apparently, special to multilevel CD methods and does not arise in corresponding unconstrained problems.
\item In the multilevel and nonlinear case, we make the CD algorithm practical by addressing efficient storage of intermediate quantities.  In particular, in Section \ref{sec:results} we show results from a full approximation storage (FAS; see \cite{Brandt1977}) implementation.  Thereby we avoid global (Newton) linearization when solving nonlinear VI problems; compare \cite{GraeserKornhuber2009}.
\end{enumerate}

Especially for the classical obstacle problem, i.e.~for a Laplacian functional, certain other methods are known to improve performance relative to the multilevel CD method \cite{GraeserKornhuber2009}.  These improved multilevel methods either track the active set or modify the nodal basis functions, and in this sense they are discrete algorithms.  By contrast the CD method applies at the level of the continuum problem (see below); it is more fundamental than the truncated monotone multigrid method \cite{Kornhuber1994}, for example, in that sense, and we can demonstrate comparable rates of convergence (Section \ref{sec:results}). % HOPE
Furthermore, truncated monotone approaches for nonlinear problems have, so far, applied a global linearization strategy \cite{GraeserKornhuber2009}, while we apply FAS (Section \ref{sec:multilevel}).  In any case, acceleration of our nonlinear multilevel CD method by such active-set and/or basis-level manipulations is a topic for further research.

>>>>

The goal of this paper is to generalize the constraint decomposition (CD) method of X.-C.~Tai \cite{Tai2003}.  The convergence of this method, for several types of decompositions, was proven by Tai for coercive variational inequality (VI) problems which arise from minimization of a convex functional over a convex set.  In its multilevel form the method has been shown to have optimal complexity for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}; see also \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.

We extend the CD method to nonlinear VI problems for which it has not been previously considered, including a new proof of convergence which removes the assumption that the problem arises from (constrained) optimization.  Furthermore we address new finite element (FE) implementation concerns in nonlinear cases, and we demonstrate the resulting performance.  The original algorithms and theory presented in \cite{Tai2003} are extended in four particular directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item We do not assume that the continuum VI problem arises from optimization of a scalar objective.  Nonetheless we can prove convergence in norm at the same rate as the original method converges in energy (Section \ref{sec:convergence}); % HOPE
examples are explored in Sections \ref{sec:vi} and \ref{sec:results}.

\item We make the observation that, if implemented appropriately, multilevel V(0,$\nu$) ``up-slash'' cycles are intrinsically more efficient than the V($\nu$,0) ``down-slash'' cycles (Section \ref{sec:multilevel}).  This observation seems to be new and should replace the default use of down-slash or even balanced (e.g.~V(1,1)) cycles; compare the comments on V(1,0) and V(1,1) cycles in \cite{GraeserKornhuber2009,Tai2003}.  A strong preference for up-smoothing is, apparently, special to multilevel CD methods and does not arise in corresponding unconstrained problems.

\item In the nonlinear case, we make the multilevel CD algorithm more practical by addressing efficient storage of intermediate quantities.  In particular, in Section \ref{sec:results} we show results from a full approximation storage (FAS; see \cite{Brandt1977}) implementation which avoids global (Newton) linearization; compare \cite{GraeserKornhuber2009}.

\item For problems of porous-media type the nonlinear operator $f$ is not coercive (Appendix A), so the existing convergence theory does not apply.  However, a full-cycle implementation of the multilevel CD method is demonstrated numerically.  The operator is approximated by a coercive operator, nonlinear in general, for the duration of the V-cycle, by ``freezing' the solution-dependent coefficient.  The resulting scheme is highly-effective for doubly-nonlinear diffusion operators (Section \ref{sec:results}).
\end{enumerate}

Note that the iterates from the CD method are always admissible, and thus the operator need only be defined on the admissible subset; Section \ref{sec:results} includes an example.  Admissible-iterate methods should allow direct solutions of certain VI problems, such as fluid-layer dynamics problems \cite{Bueler2021conservation,JouvetBueler2012}, for which non-admissible methods, such as semi-smooth methods \cite{BensonMunson2006}, require unnatural modifications of the operator formula.  The full-cycle scheme in \emph{(iv)} above, which applies to non-coercive VIs, is helpful in these problems.

Especially for the classical obstacle problem with a Laplacian operator, certain other multilevel methods are known to improve performance relative to the multilevel CD method \cite{GraeserKornhuber2009}.  These improved methods either track the active set in the discretization or modify the nodal basis functions, and in this sense they are discrete algorithms.  By contrast the CD method applies at the level of the continuum problem (Sections \ref{sec:cd} and \ref{sec:convergence}), and thus it is a more fundamental method than the truncated monotone multigrid method \cite{Kornhuber1994}, for example, and in any case we can demonstrate comparable rates of convergence (Section \ref{sec:results}). % HOPE
Truncated monotone approaches for nonlinear problems have, so far, applied a global linearization strategy \cite{GraeserKornhuber2009}, while we apply FAS (Section \ref{sec:multilevel}).  Acceleration of our nonlinear multilevel CD method via active-set and/or basis-level manipulations is a topic for further research.
